---
title: "NicheNet analysis for BMEN 4480 Final Project"
output: html_notebook
---

```{r}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
suppressMessages(library(nichenetr))
suppressMessages(library(dplyr))
suppressMessages(library(tidyverse))
suppressMessages(library(Seurat))
suppressMessages(library(ggplot2))
```


```{r}
print(getwd()) # double-check the project path is correct
data_path <- "dataset/"
```

# Load NicheNet prior models & matrices, Load dataset
### Load Nichenet models
```{r}
lr_network <- readRDS(paste0(data_path, "lr_network.rds"))
weighted_networks <- readRDS(paste0(data_path, "weighted_networks.rds"))
ligand_target_matrix <- readRDS(paste0(data_path, "ligand_target_matrix.rds"))
```

```{r}
lr_network[1:5, 1:length(lr_network)]
```

```{r}
ligand_target_matrix[1:5, 1:5]
```

```{r}
prior_models <- list(lr_network=lr_network, weighted_networks=weighted_networks, ligand_target_matrix=ligand_target_matrix)
```



### Load universal expressed genes and cluster-wise specific "marker genes" top DEGs
```{r}
expr_genes <- read.csv(paste0(data_path, "expressed_genes.csv"), header = 1)
marker_genes <- read.csv("dataset/cluster_marker_genes.csv", header = 1)
```

```{r}
expr_genes[1:ncol(expr_genes)]
```
```{r}
marker_genes[1:ncol(marker_genes)]
```

### Load count matrix (filtered, log-normal transformed and batch-corrected)
```{r}
counts <- read.csv(paste0(data_path, "batchnorm_counts.csv"), row.names = "Cell_ID", header = 1)
```

```{r}
cluster_assignment <- counts$Cluster_ID
counts$Cluster_ID <- NULL
counts[1:5, 1:5]
```

### Create Seurat object for the count matrix, add cluster assignment as the metadata
```{r}
counts_s <- CreateSeuratObject(counts = t(counts), project = "raw")
counts_s <- AddMetaData(
  object = counts_s,
  metadata = cluster_assignment,
  col.name = "cluster.ident"
)
```

# Running NicheNet
### Import helper function
```{r}
require(gridExtra)
suppressMessages(library(hash))
source("rl_utils.R")
```

```{r}
title <- "Macrophage2"
```

```{r}

```

### Wrapper functions
```{r}
#' Wrapper function to run a single NicheNet analysis (Sender cluster -> Receiver cluster)
#' @param counts_s Gene by Cell count matrix in Seurat object
#' @param sender Name for sender cluster
#' @param receiver Name for receiver cluster
#' @param prior_models NicheNet prior networks
#' @return list of prediction raw results & plots
runNicheNet <- function(sender, receiver, counts_s, prior_models, fig_path = "/plots/") { 
  print(paste("Predicting RL interactions from", sender, "-->", receiver, sep = " "))
  
  sc_expr_genes <- expr_genes[[sender]]
  rc_expr_genes <- expr_genes[[receiver]]
  rc_marker_gsoi <- marker_genes[[receiver]] # genes of inteset (DEGs / marker genes) for the receiver cluster
  
  result_list <- get_RL_pred(count_matrix = counts_s,
                             geneset_oi = rc_marker_gsoi,
                             es = sc_expr_genes,
                             er = rc_expr_genes,
                             from_count = FALSE,
                             N = 25,
                             prior_models = prior_models,
                             name_l = sender,
                             name_r = receiver,
                             fig_path = fig_path)
  
    # visualization
    p1 <- result_list$p_ligand_receptor_network
    p2 <- DotPlot(counts_s,
                  features = result_list$best_upstream_ligands %>% rev(),
                  cols = "RdBu",
                  group.by = "cluster.ident"
                 ) + RotatedAxis() + xlab(paste("Ligand in", sender))
    p3 <- DotPlot(counts_s,
                  features = result_list$top_receptors %>% rev(),
                  cols = "RdBu",
                  group.by = "cluster.ident"
                 ) + RotatedAxis() + xlab(paste("Receptors in", receiver)) + theme(
                                        axis.text.x=element_text(angle=45,size = rel(0.5), margin = margin(0.2, unit = "cm"), vjust = 1))
    
    res <- list(pred_lr_network=result_list$pred_lr_network,
                top_ligands=result_list$best_upstream_ligands, 
                top_receptors=result_list$top_receptors, 
                fig=list(lr_network=p1, dot_ligands=p2, dot_receptors=p3))
    
    return (res)
}
```


```{r}
# Wrapper function to save plots
saveNicheNetPlots <- function(fig_list, sender, receiver, fig_path = "/plots/") {
    fig_path = paste0(getwd(), fig_path)
    if (!dir.exists(fig_path)) {
        dir.create(fig_path)
    }
    
    tiff(filename = paste0(fig_path, "rl_", sender, '_', receiver, ".tiff"), width = 10, height = 6, units = "in", res = 300)
    print(fig_list$lr_network)
    dev.off()

    tiff(filename = paste0(fig_path, "ligand_", sender, '_', receiver, ".tiff"), width = 8, height = 5, units = "in", res = 300)
    print(fig_list$dot_ligands)
    dev.off()
    
    tiff(filename = paste0(fig_path, "receptor_", sender, '_', receiver, ".tiff"), width = 8, height = 5, units = "in", res = 300)
    print(fig_list$dot_receptors)
    dev.off()
}
```


### Visualize NicheNet RL interction results for tumor & macrophage pair
```{r}
#' Wrapper function to run a single NicheNet analysis (Sender cluster -> Receiver cluster)
#' @param counts_s Gene by Cell count matrix in Seurat object
#' @param sender Name for sender cluster
#' @param receiver Name for receiver cluster
#' @param prior_models NicheNet prior networks
#' @return list of prediction raw results & plots
```

```{r}
tumor_clusters <- c("Transformed1", "Transformed2")
macrophage_clusters <- c("Macrophage1", "Macrophage2")

pred_lr_network_list = list()

for (tumor in tumor_clusters) {
  for (macrophage in macrophage_clusters) {
    
    # Ligand --> Receptor: Tumor cluster --> Macrophage
    res_tumor_to_macrophage <- runNicheNet(
      counts_s = counts_s,
      sender = tumor,
      receiver = macrophage,
      prior_models = prior_models
    )
    pred_lr_network_list[[paste(tumor, macrophage, sep = '_')]] <- res_tumor_to_macrophage$pred_lr_network
    saveNicheNetPlots(res_tumor_to_macrophage$fig, sender = tumor, receiver = macrophage)
    
    # Ligand --> Receptor: Macrophage --> Tumor cluster
    res_macrophage_to_tumor <- runNicheNet(
      counts_s = counts_s,
      sender = macrophage,
      receiver = tumor,
      prior_models = prior_models
    )
    pred_lr_network_list[[paste(macrophage, tumor, sep = '_')]] <- res_macrophage_to_tumor$pred_lr_network
    saveNicheNetPlots(res_macrophage_to_tumor$fig, sender = macrophage, receiver = tumor)
    gc()
    
  }
}
```


# Save top R-L interactions predicted from each tumor-macrophage cluster
```{r}
# Helper function
topPredRLscores <- function(pred_lr_network, n = 50) {
    top_prior_scores <- rev(sort(unlist(pred_lr_network)))
    top_prior_scores <- top_prior_scores[top_prior_scores > 0][1:n]
    indices <- lapply(top_prior_scores, function(x) which(pred_lr_network == x, arr.ind = TRUE))
    
    top_score_df <- data.frame(Ligand = character(), Receptor = character(), Score = numeric())
    for (i in 1:n) {
        top_score_df <- rbind(top_score_df, data.frame(
            Ligand = colnames(pred_lr_network)[indices[[i]][2]],
            Receptor = rownames(pred_lr_network)[indices[[i]][1]],
            Score = top_prior_scores[i]
        ))
    }
    
    return(top_score_df)
}
```


### Transformed 1 --> Macrophage 1
```{r}
topPredRLscores(pred_lr_network_list$Transformed1_Macrophage1)
```

### Macrophage 1 --> Transformed 1
```{r}
topPredRLscores(pred_lr_network_list$Macrophage1_Transformed1)
```

### Transformed 2 --> Macrophage 1
```{r}
topPredRLscores(pred_lr_network_list$Transformed2_Macrophage1)
```

### Macrophage 1 --> Transformed 2
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed1)
```

### Transformed 1 --> Macrophage 2
```{r}
topPredRLscores(pred_lr_network_list$Transformed1_Macrophage2)
```

### Macrophage 2 --> Transformed 1
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed1)
```

### Transformed 2 --> Macrophage 2
```{r}
topPredRLscores(pred_lr_network_list$Transformed2_Macrophage2)
```

### Macrophage 2 --> Transformed 2
```{r}
topPredRLscores(pred_lr_network_list$Macrophage2_Transformed2)
```

### Save top 50 R-L pairs to csv
```{r}
output_path <- "./top_rl_pairs/"
for (tumor in tumor_clusters) {
  for (macrophage in macrophage_clusters) {
    keyword1 <- paste(tumor, macrophage, sep = "_")
    keyword2 <- paste(macrophage, tumor, sep = "_")
    
    write.table(data.frame(topPredRLscores(pred_lr_network_list[[keyword1]])), 
                paste0(output_path, keyword1, ".csv"), sep=',' )
    write.table(data.frame(topPredRLscores(pred_lr_network_list[[keyword2]])), 
                paste0(output_path, keyword2, ".csv"), sep=',' )
  }
}
```
